---
title: "Athens Property Prices with BARTSIMP"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Athens Property Prices with BARTSIMP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

This vignette demonstrates a small spatial regression workflow with
`BARTSIMP` using the `properties` point data from the `spData` package.
The data record property prices in Athens, Greece. We model price per square
meter (`prpsqm`) using property age (`age`) and distance to the nearest metro
station (`dist_metro`), while using the point geometry as spatial coordinates.

```{r setup}
library(BARTSIMP)
library(sf)
library(spData)
```

## Data Preparation

The `properties` object is an `sf` point data set. Its coordinate reference
system is GGRS87 / Greek Grid, so the point coordinates are already projected
in meters.

```{r data}
data("properties", package = "spData")

properties_attr <- sf::st_drop_geometry(properties)
properties_small <- properties[
  complete.cases(properties_attr[, c("prpsqm", "age", "dist_metro")]),
]

coords <- sf::st_coordinates(properties_small)

athens_df <- data.frame(
  prpsqm = properties_small$prpsqm,
  age = properties_small$age,
  dist_metro = properties_small$dist_metro,
  s1 = coords[, 1],
  s2 = coords[, 2]
)

head(athens_df)
summary(athens_df[, c("prpsqm", "age", "dist_metro")])
```

For a quick demonstration, use a reproducible train/test split. The predictors
are scaled before fitting so the two covariates have comparable numeric ranges.

```{r split}
set.seed(2026)

n_train <- min(180, floor(0.7 * nrow(athens_df)))
train_id <- sample(seq_len(nrow(athens_df)), n_train)
test_id <- setdiff(seq_len(nrow(athens_df)), train_id)

test_id <- test_id[seq_len(min(60, length(test_id)))]

x_all <- scale(athens_df[, c("age", "dist_metro")])
x_all <- as.data.frame(x_all)

x_train <- x_all[train_id, , drop = FALSE]
y_train <- athens_df$prpsqm[train_id]
s_train <- athens_df[train_id, c("s1", "s2")]

x_test <- x_all[test_id, , drop = FALSE]
y_test <- athens_df$prpsqm[test_id]
s_test <- athens_df[test_id, c("s1", "s2")]
```

The spatial distribution of the selected training and test points can be
checked before fitting.

```{r plot-split}
plot(
  athens_df$s1,
  athens_df$s2,
  col = "grey80",
  pch = 16,
  asp = 1,
  xlab = "Greek Grid easting",
  ylab = "Greek Grid northing"
)
points(s_train$s1, s_train$s2, col = "#0072B2", pch = 16)
points(s_test$s1, s_test$s2, col = "#D55E00", pch = 17)
legend(
  "topright",
  legend = c("Training", "Test"),
  col = c("#0072B2", "#D55E00"),
  pch = c(16, 17),
  bty = "n"
)
```

## Fit BARTSIMP

The following code fits `BARTSIMP` to the Athens property data. The iteration
counts are intentionally small for demonstration. For analysis, increase
`ntree`, `ndpost`, `nskip`, and `nwarmup`, then inspect convergence and
posterior predictive summaries.

```{r fit, eval = FALSE}
fit <- bartsimp(
  x.train = x_train,
  y.train = y_train,
  x.test = x_test,
  s1 = s_train$s1,
  s2 = s_train$s2,
  s1.test = s_test$s1,
  s2.test = s_test$s2,
  ntree = 50,
  ndpost = 100,
  nskip = 50,
  nwarmup = 100,
  seed = 2026,
  usecoords = TRUE,
  range_new = 2500,
  sig_m_new = sd(y_train) / 2,
  sigest_new = sd(y_train) / 2
)
```

The posterior mean of the BART component for the test set can be compared with
the observed test outcomes.

```{r bart-summary, eval = FALSE}
bart_pred <- colMeans(fit$yhat.test)

sqrt(mean((bart_pred - y_test)^2))
plot(
  y_test,
  bart_pred,
  xlab = "Observed price per square meter",
  ylab = "Predicted price per square meter",
  pch = 16
)
abline(0, 1, col = "grey40", lty = 2)
```

## Add Spatial Residual Correction

`bartsimp_predict()` combines posterior BART draws with a Matern spatial field
fit to the training residuals. It uses the mesh returned by `bartsimp()`.

```{r spatial-predict, eval = FALSE}
spatial_pred <- bartsimp_predict(
  y_train = y_train,
  tree_draws_train = fit$yhat.train,
  tree_draws_test = fit$yhat.test,
  s_train = s_train,
  s_test = s_test,
  kappas = fit$kappas,
  sigmams = fit$sigmams,
  sigma_obs = fit$sigma,
  mesh = fit$mesh,
  verbose = TRUE
)

pred_mean <- colMeans(spatial_pred$prediction_mean_test)

sqrt(mean((pred_mean - y_test)^2))
```

The posterior draws can also be summarized into pointwise uncertainty
intervals.

```{r intervals, eval = FALSE}
pred_interval <- apply(
  spatial_pred$prediction_mean_test,
  2,
  quantile,
  probs = c(0.05, 0.5, 0.95)
)

t(pred_interval[, seq_len(min(6, ncol(pred_interval)))])
```
